---
title: "Exploratory Data Analysis | Uber Trips on Easter Day (NYC)"
format: html
---
Code written in R-Studio (stat20.datahub.berkeley.edu), and transferred to GitHub file
Worktime: 18 hours

Observational unit: one Uber trip in NYC on Easter 2014
Variables (within original apr dataframe):
  Date/Time : The date and time of the Uber pickup 
  Lat : the latitude of the pickup location  
  Lon : the longitude of the pickup location 
  Base : the TLC base code associated with the driver

Installing necessary libraries/packages:
```{r}
library(tidyverse)
library(ggplot2)
library(ggthemes)
library(dplyr)
library(readr)
library(patchwork)
```

Loading the raw dataset from Github user harshr28:
```{r}
#Github: harshr28, uber-raw-data-apr14.csv
apr <- read.csv("https://raw.githubusercontent.com/harshr28/ubereda/main/uber-raw-data-apr14.csv")
```

Data wrangling process:
```{r}
#Filter Uber Trips for only Easter using filter()
new_df <- apr %>%
  filter(Date.Time > "4/20/2014 0:00:00", 
         Date.Time < "4/20/2014 23:59:00") 
  
#Henry ~ https://stackoverflow.com/questions/8996134/extract-vectors-from-strsplit-list-without-using-a-loop
#Use strsplit() to convert Date.Time vector into list formatted [i][1]"m/d/y" and [i][2]"hr:min:sec" for i in range 1 to length(new_df)
#Unlist the splitted vector using unlist() then indexing by the date and time items within each nested list 
splitted <- strsplit(as.character(new_df$Date.Time), " ")
part1 <- unlist(splitted)[2*(1:length(new_df$Date.Time))-1]
part2 <- unlist(splitted)[2*(1:length(new_df$Date.Time))]

#Create new dataframe that includes the temporal vectors initialized above, along with the remaining columns in the new_df dataframe
Day <- c(part1)
Time <- c(part2)
new_df <- data.frame(Day,Time,new_df$Lat,new_df$Lon,new_df$Base)

#Sort time vector using arrange() to observe inherent ordering and record correct indices accordingly in a vector (military time)
new_df <- arrange(new_df, Time)
index_vector <- c(1:536, 6930:7362, 9223:9497, 537:1624, 1625:4951, 4952:6929, 7363:7993, 7994:9222)

#In the order specified by index_vector, done in recursive fashion, append the index ranges of new_df to a new vector using rbind() 
#Save the final (or theoretically, the seventh) vector to the new_df dataframe
first <- rbind(new_df[1:536, ], new_df[6930:7362, ])
second <- rbind(first, new_df[9223:9497, ])
third <- rbind(second, new_df[537:1624, ])
fourth <- rbind(third, new_df[1625:4951, ])
fifth <- rbind(fourth, new_df[4952:6929, ])
sixth <- rbind(fifth, new_df[7363:7993, ])
new_df <- rbind(sixth, new_df[7994:9222, ])

#Manually collapse time vector using using as.character() within fct_collapse(), then create an Index column from 1 to length(new_df)
#Create four levels: morning (12:00am - 11:59am), afternoon (12:00pm - 4:59pm), Evening (5:00pm - 8:45pm), Night (9:00pm - 11:59pm) 
new_df <- new_df %>%
  mutate(Index = c(1:9497),
         time_category = fct_collapse(as.character(Index),
                                      "Morning" = 1:1624,
                                      "Afternoon" = 1625:4951,
                                      "Evening" = 4952:7993,
                                      "Night" = 7994:9497))

#Utilize ordered factor() to obtain desired arrangement of column
new_df$time_category <- factor(new_df$time_category, 
                         levels = c("Morning","Afternoon","Evening","Night"))

#Mutate logical vector with condition with ifelse() to determine if a trip is from Brooklyn, where most trips tend to be made from
new_df <- new_df %>%
  mutate(long_west = ifelse(new_df.Lat > 40.6 & new_df.Lat < 40.8 & new_df.Lon < -73.8 & new_df.Lon > -74.05, TRUE, FALSE))
```

Visualizations:
1.
```{r}
#Bar plot for trips by time category (vector(s): time_category)
tod_group <- new_df %>%
  ggplot(aes(x = time_category)) + 
  geom_bar(fill = "skyblue", color = "black") + 
  labs(x = "Time of Day",
       y = "Number of Trips",
       title = "Trips by Time of Day on Easter (NYC)") +
  theme_classic()
tod_group
```
2.
```{r}
#Density of trips throughout Easter in NYC (vector(s): Time, date_time_format(modified vector)
#WARNING: the date_time_format vector is formatted as [2023-12-20 hr:min:sec] after being converted into as.POSIXct class
annotate_1 <- as.POSIXct("00:00:00", format = "%H:%M:%S")
annotate_2 <- as.POSIXct("12:00:00", format = "%H:%M:%S")
annotate_3 <- as.POSIXct("17:00:00", format = "%H:%M:%S")
annotate_4 <- as.POSIXct("21:00:00", format = "%H:%M:%S")
annotate_5 <- as.POSIXct("24:00:00", format = "%H:%M:%S")

time_density <- new_df %>%
  mutate(date_time_format = as.POSIXct(new_df$Time, format = "%H:%M:%S")) %>%
  ggplot(aes(x = date_time_format)) + 
  geom_density(fill = "dodgerblue4", alpha = 0.9) +
  scale_x_datetime(labels = scales::time_format("%H:%M:%S")) +
  labs(x = "Military Time",
       y = "Density") +
  geom_vline(xintercept = as.numeric(annotate_1), linetype = "dashed", color = "lightblue", size = 1) +
  geom_vline(xintercept = as.numeric(annotate_2), linetype = "dashed", color = "lightblue", size = 1) +
  geom_vline(xintercept = as.numeric(annotate_3), linetype = "dashed", color = "lightblue", size = 1) +
  geom_vline(xintercept = as.numeric(annotate_4), linetype = "dashed", color = "lightblue", size = 1) +
  geom_vline(xintercept = as.numeric(annotate_5), linetype = "dashed", color = "lightblue", size = 1) +
  theme_classic() 
time_density
```
3.
```{r}
#Displaying counts for time category given base (vector(s): new_df.Base, time_category)
bases_group <- new_df %>%
  ggplot(aes(x = new_df.Base, 
             fill = time_category)) + 
  geom_bar(color = "black", alpha = 0.7) + 
  labs(x = "Base",
       y = "Number of Trips",
       title = "Trips by Base on Easter (NYC)") +
  scale_fill_manual(name = "Time of Day", 
                    values = c("tan", "plum", "orange", "dodgerblue")) +
  theme_classic()

#Normalized stacked bar plot for proportions and association
bases_assoc <-  new_df %>%
  ggplot(aes(x = new_df.Base, 
             fill = time_category)) + 
  geom_bar(color = "black", position = "fill", alpha = 0.7) + 
  labs(x = "Base",
       y = "Number of Trips",
       title = "Trips by Base on Easter (NYC)") +
  scale_fill_manual(name = "Time of Day", 
                    values = c("tan", "plum", "orange", "dodgerblue")) +
  theme_classic()
bases_group / bases_assoc
```
4.
```{r}
#Maximum and minimum lattitude and longitude for filtered new_df dataframe
max_lat <- max(new_df$new_df.Lat)
min_lat <- min(new_df$new_df.Lat)
max_lon <- max(new_df$new_df.Lon)
min_lon <- min(new_df$new_df.Lon)

#Scatter plot for location of trips by time category (vector(s): new_df.lat, new_df.lon, and time_category)
location_group <- new_df %>%
  ggplot(aes(x = new_df.Lon, 
             y = new_df.Lat, 
             color = time_category)) + 
  geom_point(alpha = .5, size = 1.2) + 
  labs(x = "Longitude",
       y = "Latitude",
       title = "Pickup Density by Time of Day on Easter (NYC)") +
  scale_color_manual(name = "Time of Day", 
                    values = c("tan", "plum", "orange", "purple")) +  
  scale_x_continuous(limits = c(min_lon, max_lon)) + 
  scale_y_continuous(limits = c(min_lat, max_lat)) + 
  annotate("segment", x = -74.05, xend = -74.05, y = 40.8, yend = 40.6) + 
  annotate("segment", x = -74.05, xend = -73.8, y = 40.6, yend = 40.6) + 
  annotate("segment", x = -73.8, xend = -73.8, y = 40.6, yend = 40.8) + 
  annotate("segment", x = -73.8, xend = -74.05, y = 40.8, yend = 40.8) + 
  annotate("segment", x = -73.8, xend = -73.65, y = 40.8, yend = 40.9) + 
  annotate("text", x = -73.6, y = 40.92, label = "Brooklyn Area") + 
  theme_clean()
location_group
```
5.
```{r}
#Saving summary statistic for the proportion of trips that were taken from Brooklyn, for each time category
prop_values <- new_df %>%
  group_by(time_category) %>%
  summarize(prop = mean(long_west))

#Displaying counts (vector(s): time_category, long_west (ordered and renamed levels))
prop_group <- new_df %>%
  ggplot(aes(x = time_category, fill = factor(long_west, levels = c(TRUE, FALSE)))) + 
  geom_bar(color = "black", alpha = 0.7) + 
  labs(x = "Time of Day",
       y = "Number of Trips",
       title = "Trips by Time of Day on Easter (NYC)") +
  scale_fill_manual(name = "Trip from Upper Brooklyn",
                    labels = c("Yes", "No"),
                    values = c("skyblue", "dodgerblue")) +
  theme_classic()

#Normalized stacked bar plot for proportions and association
prop_assoc <- new_df %>%
  ggplot(aes(x = time_category, fill = factor(long_west, levels = c(TRUE, FALSE)))) + 
  geom_bar(position = "fill", color = "black", alpha = 0.7) + 
  labs(x = "Time of Day",
       y = "Number of Trips",
       title = "Trips by Time of Day on Easter (NYC)") +
  scale_fill_manual(name = "Trip from Upper Brooklyn",
                    labels = c("Yes", "No"),
                    values = c("skyblue", "dodgerblue")) +
  theme_classic()
prop_values
prop_group / prop_assoc 
```
